\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-08-01 mers-structure.tex
High resolution figures, some text changes.

The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we estimate a median of 56 separate camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of \tbc{XXX (95\% HPD: XXX--XXX)} and human viruses showing a scaled effective population of \tbc{XXX (95\% HPD: XXX--XXX)}.

@@

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from 174 human viruses and 100 camel viruses.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
	The vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
  \tbc{The previous two sentences (``Some of the...'') seem unnecessary and confusingly written. I would just drop them.}
	While large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
	}
	\label{mcc}
\end{figure}

@@

% \begin{figure}[h]
% \centering
%
% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
% 	}
% 	\label{coocurrence}
% \end{figure}
\begin{figure}%
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to \tbc{XXX}.
    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
  	While some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels, adding support to the idea that camels are often co-infected. %\lmc{I'm sure this is important, but I didn't get why. Referees might not either...}
    \tbc{I would favor just dropping the ``While some of apparent...'' sentence. Confusing and I don't see what it adds.}
  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
    }
    \label{recombination}
\end{figure}

@@

\subsection*{MERS-CoV epidemiology}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
\tbc{This seems way overly broad to me. I much preferred the Introductions take on genomic epi revealing important population structure. Just revise to keep this focused on pathogen transmission. Identification of reservoirs, source-sink, all very important.}
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
While there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as $R_0$ or the number of zoonotic introductions of MERS-CoV into humans.
\tbc{This sentence (``While there...'') is overly like Intro.}

@@

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.

\tbc{Perhaps close with discussion of evolution of host tropism? So far outbreaks have been self-limiting, but continued circulation in camels and continued spill-over into humans provides continued chances for evolution of a more human transmissible virus.}

>>>>

The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we estimate a median of 56 separate camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of 3.49 (95\% HPD: 2.71--4.38) and human viruses showing a scaled effective population of 0.24 (95\% HPD: 0.14--0.34).

@@

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from 174 human viruses and 100 camel viruses.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
	The vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
	While large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
	}
	\label{mcc}
\end{figure}

@@

% \begin{figure}[h]
% \centering
%
% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
% 	}
% 	\label{coocurrence}
% \end{figure}
\begin{figure}%
    \centering
    \includegraphics[width=0.95\textwidth]{figures/mers_recombinant_features.png}
    \caption{\textbf{Recombinant features of MERS-CoV phylogenies.}
    A) Heatmap showing the posterior probability that viruses from trees of different genomic fragments belong to the same introduction event.
    Genomic fragment 1 represents nucleotide positions 1 to 21000 and genomic fragment 2 represents nucleotide positions 21001 to \tbc{XXX}.
    The same set of viruses are arrayed on the $x$-axis and on the $y$-axis; the $x$-axis shows identity of these viruses along genomic fragment 1 and the $y$-axis shows identity of these viruses along genomic fragment 2.
    Viruses are ordered by their appearance in tree of genomic fragment 2 reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
    B) Phylogenies derived from genomic fragment 1 (left) and genomic fragment 2 (right), reduced to just the human tips.
  	Identical taxa are connected via coloured lines with colour indicating vertical position in the tree on the right.
  	Branches are coloured by inferred ancestral host state (human in blue, camel in orange).
  	Human clusters in blue change phylogenetic positions between the trees together, with little incongruence within clusters.
  	This is evidence for recombinant viruses generated in the camel reservoir entering human populations.
    }
    \label{recombination}
\end{figure}

@@

\subsection*{MERS-CoV epidemiology}
%\lmc{Detecting?}
Understanding population structure yields important insight into any biological system.
It also often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
\tbc{This seems way overly broad to me. I much preferred the Introductions take on genomic epi revealing important population structure. Just revise to keep this focused on pathogen transmission. Identification of reservoirs, source-sink, all very important.}
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula by using sequence data collected from both hosts (174 from humans and 100 from camels).

@@

MERS-CoV may join the list of pathogens able to jump species barriers but not spread efficiently in the new host, but every system is distinct.
Pathogens such as \textit{Bacillus anthracis}, Andes hantavirus \citep{martinez_person--person_2005}, monkeypox \citep{reed_detection_2004}, triple reassortant and H3N2v influenza A viruses \citep{shinde_triple-reassortant_2009,epperson_human_2013} belong to this list and yet only triple reassortant viruses eventually contributed to a pandemic, due to epidemiology and phylodynamics confined to influenza A virus.
When information about a system is lacking either because it is recent or entirely novel or if the system is inherently unpredictable, sequence data can and should be used jointly with whatever data sources are available to provide the unique pathogen perspective on an infectious disease outbreak.
Although MERS-CoV seems to cause self-limiting outbreaks in humans with no evidence of worsening outbreaks over time, sustained evolution of the virus in the reservoir and flow of viral lineages into humans does not eliminate the possibility that a more transmissible variant of MERS-CoV will emerge.
We encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases.
